---
title: "ESFS Workflow with Seurat"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy    = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  eval    = FALSE
)
```

> **ESFS — Entropy Sorting Feature Selection**
> Radley A. *et al.* (2026)
> Preprint: <https://www.biorxiv.org/content/10.64898/2026.01.26.701684v1>
> Source: <https://github.com/aradley/ESFS>

---

## Overview

ESFS (Entropy Sorting Feature Selection) identifies marker genes and performs
feature selection in single-cell RNA-seq data using information-theoretic
metrics.  Unlike PCA-based workflows it operates directly in gene expression
space, preserving biological interpretability and avoiding assumptions about
cell-type structure.

This vignette demonstrates the complete ESFS workflow using **SeuratESFS**,
an R wrapper that calls the ESFS Python package under the hood via
[reticulate](https://rstudio.github.io/reticulate/).

## Installation

### Step 1 — Install SeuratESFS

```{r install-r-pkg}
remotes::install_github("aradley/SeuratESFS")
```

### Step 2 — Set up the Python environment (once per machine)

```{r install-python-env}
library(SeuratESFS)

# Auto-detects hardware:
#   Apple Silicon → MLX/Metal backend
#   NVIDIA GPU    → CUDA/CuPy backend
#   Otherwise     → CPU/Numba backend
install_esfs()
```

After installation you will be prompted to add one line to `~/.Rprofile`:

```r
options(esfs.conda_env = "r-esfs")
```

This ensures the correct Python environment is loaded automatically in every
future R session.

### Step 3 — Verify the installation

```{r verify}
check_esfs()
```

---

## Loading packages

**Important:** load `SeuratESFS` *before* Seurat (or any other package that
may initialise Python) so the `.onLoad` hook can configure reticulate before
Python starts.

```{r load-pkgs}
library(SeuratESFS)
library(Seurat)
```

---

## Example data

We use the `pbmc3k` dataset from
[SeuratData](https://github.com/satijalab/seurat-data), which contains 2,700
peripheral blood mononuclear cells.

```{r load-data}
# install.packages("remotes")
# remotes::install_github("satijalab/seurat-data")
library(SeuratData)
InstallData("pbmc3k")
data("pbmc3k")

# Standard Seurat pre-processing
pbmc3k <- NormalizeData(pbmc3k)
pbmc3k <- FindVariableFeatures(pbmc3k, nfeatures = 3000)
pbmc3k <- ScaleData(pbmc3k)
pbmc3k <- RunPCA(pbmc3k)
pbmc3k <- FindNeighbors(pbmc3k, dims = 1:10)
pbmc3k <- FindClusters(pbmc3k, resolution = 0.5)
pbmc3k <- RunUMAP(pbmc3k, dims = 1:10)
```

---

## ESFS workflow

### Step 1 — Check the active backend

```{r backend-info}
GetESFSBackendInfo()
# Example output on Apple Silicon:
# "Backend: MLX/Metal (Apple Silicon GPU), Precision: float32"
```

### Step 2 — Scale the expression matrix

`CreateESFSScaledMatrix()` scales each gene to \[0, 1\] using per-gene
percentile clipping and removes genes with very low total expression.

```{r scale-matrix}
pbmc3k <- CreateESFSScaledMatrix(
  pbmc3k,
  clip_percentile = 97.5,
  min_total_expr  = 50L
)
```

### Step 3 — Calculate pairwise ES matrices (ES-GSS)

This is the computationally intensive step. It calculates four pairwise
information-theoretic metrics (ESS, EP, SW, SG) between all gene pairs.

GPU acceleration is used automatically when available. On Apple Silicon the
Metal GPU is used; on NVIDIA hardware CUDA/CuPy is used; otherwise Numba
JIT-compiled CPU code runs in parallel.

*Note: the first call incurs a one-time Numba compilation delay (~10–30 s)
that is cached for subsequent sessions.*

```{r calc-es-matrices}
pbmc3k <- CalcESMatrices(
  pbmc3k,
  label         = "Self",   # compare each gene to every other gene
  save_matrices = c("ESSs", "EPs"),
  n_cores       = -1L       # use all available CPU cores
)
```

### Step 4 — Rank genes by network connectivity

Builds a weighted gene network from the ES matrices and ranks genes by
normalised connectivity.

```{r rank-genes}
pbmc3k <- ESRankGenes(
  pbmc3k,
  label         = "Self",
  ep_threshold  = 0.0,
  ess_threshold = 0.01,
  min_edges     = 5L
)
```

### Step 5 — Visualise the top-ranked genes in a UMAP

Embeds the top-ranked genes in a 2-D UMAP based on their pairwise ESS
correlation matrix.  Clustering the gene UMAP reveals groups of co-expressed
genes.

```{r gene-umap}
gene_umap <- PlotESFSGeneUMAP(
  pbmc3k,
  top_ranked_genes = 3000,
  clustering       = "hdbscan",   # or an integer for KMeans
  umap_min_dist    = 0.1,
  umap_neighbours  = 20L
)

# gene_umap$genes     — names of the plotted genes
# gene_umap$labels    — cluster label for each gene
# gene_umap$embedding — UMAP coordinates (n_genes × 2)
```

### Step 6 — Compute cell UMAPs for each gene cluster

For each cluster of genes identified above, project all cells into a 2-D UMAP
using that cluster's gene set as features.

```{r cell-umaps}
cell_umap_result <- GetESFSCellUMAPs(
  pbmc3k,
  gene_clust_labels = gene_umap$labels,
  top_ess_genes     = gene_umap$genes,
  n_neighbors       = 30L,
  min_dist          = 0.1,
  log_transformed   = TRUE,
  return_model      = FALSE
)

embeddings    <- cell_umap_result[[1]]   # list of embeddings, one per cluster
selected_genes <- cell_umap_result[[2]] # list of gene sets, one per cluster
```

### Step 7 — Plot cell UMAPs coloured by cell type

```{r plot-cell-umaps-label}
# Colour by Seurat cluster identity
pbmc3k$seurat_clusters_char <- as.character(pbmc3k$seurat_clusters)

PlotESFSCellUMAPs(
  pbmc3k,
  gene_cluster_embeddings    = embeddings,
  gene_cluster_selected_genes = selected_genes,
  cell_label = "seurat_clusters_char",
  ncol       = 2L
)
```

```{r plot-cell-umaps-gene}
# Colour by a specific gene's expression
PlotESFSCellUMAPs(
  pbmc3k,
  gene_cluster_embeddings    = embeddings,
  gene_cluster_selected_genes = selected_genes,
  cell_label           = "CD3D",
  log2_gene_expression = TRUE,
  ncol                 = 2L
)
```

---

## Optional: Select N non-redundant marker genes (ES-FMG)

ES-FMG uses simulated annealing to find a compact set of genes that captures
maximally distinct expression patterns.

```{r es-fmg}
fmg_result <- RunESFMG(
  pbmc3k,
  N           = 100L,
  label       = "Self",
  num_reheats = 3L,
  resolution  = 1.0    # 1 = maximally distinct; <0.5 for trajectories
)

# fmg_result$genes   — character vector of selected marker genes
# fmg_result$scores  — numeric ES-FMG score for each gene
# fmg_result$object  — updated Seurat object
pbmc3k <- fmg_result$object
head(fmg_result$genes)
```

---

## Optional: KNN expression smoothing

Smooth gene expression by averaging each cell over its k nearest neighbours
in the correlation distance space defined by the top ESFS genes.

```{r knn-smooth}
pbmc3k <- KNNSmoothESFS(
  pbmc3k,
  use_genes    = gene_umap$genes,   # top-ranked genes define the distance metric
  knn          = 30L,
  log_scale    = FALSE,
  max_size_gb  = 2.0   # extract smoothed matrix to R only if < 2 GB
)

# For small datasets the smoothed expression is available as a Seurat layer:
# FeaturePlot(pbmc3k, features = "CD3D", layer = "esfs.smoothed")
```

---

## Optional: Save and reload UMAP models

UMAP models can be saved to disk and reloaded to project new cells into
existing embeddings.

```{r save-model}
# Re-run with return_model = TRUE to get saveable models
cell_umap_models <- GetESFSCellUMAPs(
  pbmc3k,
  gene_clust_labels = gene_umap$labels,
  top_ess_genes     = gene_umap$genes,
  return_model      = TRUE
)

model_list  <- cell_umap_models[[1]]
gene_sets   <- cell_umap_models[[2]]

# Save the model for cluster 0
SaveESFSUMAPModel(
  model     = model_list[[1]],
  gene_list = gene_sets[[1]],
  filepath  = "cluster0_umap.joblib"
)

# Reload later
bundle <- LoadESFSUMAPModel("cluster0_umap.joblib")
# bundle$model       — fitted umap-learn UMAP model (supports .transform())
# bundle$gene_list   — genes used for the embedding
# bundle$log_transformed — whether log2(X+1) was applied
```

---

## Backend control

```{r backend-control}
# Check current backend
GetESFSBackendInfo()

# Force a specific backend (takes effect for all subsequent ESFS calls)
UseESFSMLX()    # Apple Silicon Metal
UseESFSGPU()    # NVIDIA CUDA
UseESFSCPU()    # CPU / Numba

# Or configure with full control
ConfigureESFSBackend(gpu = TRUE, upcast = FALSE, verbose = TRUE)
```

**Backend capabilities:**

| Operation | CUDA (NVIDIA) | MLX (Apple Silicon) | CPU |
|-----------|:---:|:---:|:---:|
| ES matrix calculations | GPU | GPU | Numba JIT |
| ES-CCF / ES-FMG | CPU | CPU | CPU |
| Cell UMAP KNN | cuML GPU | CPU | CPU |
| Cell UMAP layout | cuML GPU | CPU | CPU |
| Gene UMAP | CPU | CPU | CPU |
| KNN smoothing (KNN step) | cuML GPU | CPU | CPU |

---

## Session info

```{r session-info}
sessionInfo()
```
