---
title: "ESFS Workflow — Peri-implantation Human Embryo Example"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy      = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message   = FALSE,
  warning   = FALSE,
  eval      = FALSE
)
```

> **Reference:** Radley A. *et al.* (2026) *Entropy Sorting Feature Selection for single-cell RNA sequencing.*
> Preprint: <https://www.biorxiv.org/content/10.64898/2026.01.26.701684v1>
> Data: <https://figshare.com/s/4e445e7fa03cc4ccd289>

---

## Overview

This notebook is an R equivalent of the Python Jupyter workflow for the peri-implantation human embryo
scRNA-seq dataset (7,732 cells × 16,633 genes, multiple labs/datasets integrated).

The workflow demonstrates:

1. Identifying a compact set of biologically informative genes (ES-GSS)
2. Removing dataset-specific batch-effect genes without smoothing or integration
3. Generating high-resolution cell UMAPs that reveal developmental trajectories
4. Finding optimal marker genes (ES-CCF + ES-FMG)
5. Projecting query data onto the reference embedding

**Important:** Load `SeuratESFS` *before* Seurat and any other package that may
initialise Python.

---

## Setup

### Data path — edit this

```{r data-path}
path <- "/Users/radleya/The Francis Crick Dropbox/BriscoeJ/Radleya/ESFS_Paper/Paper_Submission/Data/Peri_implantation_Human_Embryo_Example/"
```

### Load packages

```{r load-packages}
# SeuratESFS MUST be loaded before Seurat to configure the Python environment first
library(SeuratESFS)
library(Seurat)
library(Matrix)
library(ggplot2)

# For reading h5ad files into Seurat — install if needed:
# BiocManager::install("zellkonverter")
library(zellkonverter)
```

### Check ESFS backend

```{r check-backend}
check_esfs()
GetESFSBackendInfo()
```

---

## Load data

```{r load-data}
# Read h5ad into a SingleCellExperiment and convert to Seurat
sce <- zellkonverter::readH5AD(file.path(path, "ES_adata.h5ad"))
obj <- SeuratObject::as.Seurat(sce, counts = "X", data = NULL)
DefaultAssay(obj) <- "RNA"

# Verify key metadata columns
message("Cells: ", ncol(obj), "   Genes: ", nrow(obj))
message("Metadata columns: ", paste(colnames(obj[[]]), collapse = ", "))
head(obj[[c("Original_Annotation", "Timepoint", "Dataset")]])
```

---

## Step 1 — Scale the count matrix

`CreateESFSScaledMatrix()` clips each gene at the 97.5th percentile and normalises to [0, 1].
Genes with very low total expression are removed (default `min_total_expr = 50`).

```{r scale-matrix}
obj <- CreateESFSScaledMatrix(obj)
message("Genes after filtering: ", nrow(obj))
```

---

## Step 2 — Calculate pairwise ES matrices (Self)

This is the main computational step. For the human embryo dataset
(~15,000 genes after filtering), expect ~15–25 minutes on a modern CPU.
Apple Silicon (MLX/Metal) or NVIDIA (CUDA) GPUs will be used automatically if available.

*Note: on first run, Numba will JIT-compile the CPU kernels — this causes a
one-time delay of ~30 seconds before progress appears.*

```{r calc-es-self}
obj <- CalcESMatrices(
  obj,
  label         = "Self",
  save_matrices = c("ESSs", "EPs"),
  n_cores       = -1L
)
```

---

## Step 3 — Identify and exclude batch-effect genes

The dataset combines cells from multiple labs. Rather than batch integration,
ESFS identifies genes whose expression is strongly associated with the *Dataset*
label and excludes them before gene ranking.

### Build the one-hot Dataset matrix

```{r batch-onehot}
dataset_labels  <- obj$Dataset
unique_datasets <- sort(unique(dataset_labels))
message("Datasets: ", paste(unique_datasets, collapse = ", "))

dataset_onehot <- Matrix::sparseMatrix(
  i    = seq_along(dataset_labels),
  j    = match(dataset_labels, unique_datasets),
  x    = 1,
  dims = c(length(dataset_labels), length(unique_datasets)),
  dimnames = list(colnames(obj), unique_datasets)
)
```

### Calculate ES metrics against Dataset labels

```{r calc-es-batch}
obj <- CalcESMatrices(
  obj,
  label                     = "Dataset_Secondary_Features",
  save_matrices             = c("ESSs", "EPs", "SGs"),
  secondary_features_matrix = dataset_onehot
)
```

### Identify batch-effect genes

```{r find-batch-genes}
batch_ess <- Seurat::Misc(obj, "esfs_Dataset_Secondary_Features_ESSs")
# batch_ess: matrix of n_genes × n_datasets — genes with any ESS > 0.01 are batch-associated
batch_gene_mask <- apply(batch_ess > 0.01, 1, any)
batch_genes     <- rownames(obj)[batch_gene_mask]
message(sprintf("Identified %d potential batch-effect genes", length(batch_genes)))
```

---

## Step 4 — Rank genes by ES network connectivity

`ESRankGenes()` builds a weighted gene network from the Self ES matrices and ranks genes
by normalised connectivity. Batch-effect genes are excluded from the network.

Because the batch-effect step (Step 3) and the Self step (Step 2) use *different*
label keys, the Self ES matrices are preserved in `misc` and restored automatically here.

```{r rank-genes}
known_genes <- c("NANOG", "SOX17", "GATA4", "PDGFRA", "PRSS3", "SPIC")

obj <- ESRankGenes(
  obj,
  label         = "Self",
  ep_threshold  = 0.0,
  ess_threshold = 0.01,
  exclude_genes = batch_genes,
  known_genes   = known_genes,
  min_edges     = 5L
)
```

---

## Step 5 — Visualise top-ranked genes in a gene-space UMAP

Genes are embedded in 2-D based on their pairwise ESS scores.
HDBSCAN groups genes into clusters that share similar expression patterns.

Adjust `top_ranked_genes` (3,000–5,000 is typical) and `clustering` to optimise
the gene UMAP for your data.

```{r gene-umap}
gene_umap <- PlotESFSGeneUMAP(
  obj,
  top_ranked_genes  = 4000L,
  label             = "Self",
  clustering        = "hdbscan",
  known_genes       = known_genes,
  umap_min_dist     = 0.1,
  umap_neighbours   = 20L,
  hdbscan_min_size  = 50L,
  dpi               = 150L
)

# The returned list gives access to the gene cluster data:
top_ess_genes    <- gene_umap$genes   # character vector of gene names
gene_clust_labels <- gene_umap$labels  # integer cluster label per gene
message(sprintf("Gene clusters: %s", paste(sort(unique(gene_clust_labels)), collapse = ", ")))
```

---

## Step 6 — Compute cell UMAPs for each gene cluster

For each gene cluster, project all cells into a 2-D UMAP using that cluster's gene
set as features.

**Backend note:** UMAP computation is GPU-accelerated on NVIDIA (cuML) only.
On Apple Silicon and CPU, umap-learn is used (still fast for ~8,000 cells).

```{r cell-umaps-all}
cell_umaps <- GetESFSCellUMAPs(
  obj,
  gene_clust_labels = gene_clust_labels,
  top_ess_genes     = top_ess_genes,
  n_neighbors       = 50L,
  min_dist          = 0.1,
  log_transformed   = TRUE,
  return_model      = FALSE
)
```

### Plot by cell annotation

```{r plot-cell-umaps-annotation}
PlotESFSCellUMAPs(
  obj,
  gene_cluster_embeddings      = cell_umaps[[1]],
  gene_cluster_selected_genes  = cell_umaps[[2]],
  cell_label           = "Original_Annotation",
  ncol                 = 3L,
  figsize              = c(16, 5),
  marker_size          = 1L,
  log2_gene_expression = TRUE,
  dpi                  = 150L
)
```

### Plot by gene expression (NANOG)

```{r plot-cell-umaps-nanog}
PlotESFSCellUMAPs(
  obj,
  gene_cluster_embeddings      = cell_umaps[[1]],
  gene_cluster_selected_genes  = cell_umaps[[2]],
  cell_label           = "NANOG",
  ncol                 = 3L,
  figsize              = c(16, 5),
  marker_size          = 1L,
  log2_gene_expression = TRUE,
  dpi                  = 150L
)
```

---

## Step 7 — Focus on the most informative gene cluster

Inspect the gene UMAPs from Step 6 and identify the cluster that best resolves
the biological structure of interest. For the human embryo data, cluster 1
captures the most distinct developmental trajectories.

```{r cell-umaps-cluster1}
cell_umaps_c1 <- GetESFSCellUMAPs(
  obj,
  gene_clust_labels = gene_clust_labels,
  top_ess_genes     = top_ess_genes,
  specific_cluster  = c(1L),    # use c() to ensure R vector -> Python list
  n_neighbors       = 50L,
  min_dist          = 0.1,
  log_transformed   = TRUE,
  return_model      = FALSE
)

# Extract the gene set that produced this embedding
es_gss_genes <- reticulate::py_to_r(cell_umaps_c1[[2]][[1]])
message(sprintf("ES-GSS gene set: %d genes", length(es_gss_genes)))
```

```{r plot-cluster1-multi}
# By cell annotation
PlotESFSCellUMAPs(
  obj,
  gene_cluster_embeddings     = cell_umaps_c1[[1]],
  gene_cluster_selected_genes = cell_umaps_c1[[2]],
  cell_label = "Original_Annotation",
  ncol = 1L, figsize = c(5, 5), marker_size = 1L
)

# By timepoint
PlotESFSCellUMAPs(
  obj,
  gene_cluster_embeddings     = cell_umaps_c1[[1]],
  gene_cluster_selected_genes = cell_umaps_c1[[2]],
  cell_label = "Timepoint",
  ncol = 1L, figsize = c(5, 5), marker_size = 1L
)

# By dataset (check batch-effect removal worked)
PlotESFSCellUMAPs(
  obj,
  gene_cluster_embeddings     = cell_umaps_c1[[1]],
  gene_cluster_selected_genes = cell_umaps_c1[[2]],
  cell_label = "Dataset",
  ncol = 1L, figsize = c(5, 5), marker_size = 1L
)
```

---

## Step 8 — KNN expression smoothing

Smooth gene expression by averaging each cell over its 50 nearest neighbours in
correlation-distance space defined by the ES-GSS gene set.

**Backend note:** KNN computation is GPU-accelerated on NVIDIA (cuML). On Apple
Silicon and CPU, chunked BLAS matrix multiplication is used (O(N²) — slower for
large datasets). The smoothed matrix is transferred to Seurat only if ≤ `max_size_gb`.

```{r knn-smooth}
obj <- KNNSmoothESFS(
  obj,
  use_genes   = es_gss_genes,
  knn         = 50L,
  log_scale   = FALSE,
  max_size_gb = 2.0
)

# If the smoothed expression was extracted (small enough dataset), use it with Seurat:
# FeaturePlot(obj, features = "NANOG", layer = "esfs.smoothed")
```

---

## Step 9 — ES-CCF: Combinatorial Cluster Finder

ES-CCF identifies the combination of clusters that maximally correlates with each
gene's expression. This requires intentional over-clustering followed by computing
ES metrics against those cluster labels.

### Step 9a — Leiden over-clustering

Use scanpy (already in the `r-esfs` Python environment) for KNN + Leiden with the
exact settings from the Python workflow (correlation distance, n_neighbors = 50,
resolution = 8).

```{r leiden-clustering}
sc <- reticulate::import("scanpy", convert = FALSE)
np <- reticulate::import("numpy", convert = FALSE)
ad <- reticulate::import("anndata", convert = FALSE)

# Build a temporary AnnData with just the ES-GSS genes, log2 transformed
log2_counts <- log2(as.matrix(Seurat::LayerData(obj[es_gss_genes, ], layer = "counts")) + 1)

sub_adata <- ad$AnnData(X = np$array(t(log2_counts), dtype = "float32"))

sc$pp$neighbors(sub_adata, n_neighbors = 50L, n_pcs = 0L, metric = "correlation")
sc$tl$leiden(sub_adata, resolution = 8.0)

leiden_labels <- reticulate::py_to_r(sub_adata$obs["leiden"])$leiden
obj$leiden_clusters <- leiden_labels
n_clusters <- length(unique(leiden_labels))
message(sprintf("Leiden produced %d clusters", n_clusters))
```

> **Note:** `sc$tl$leiden()` requires the `leidenalg` Python package. If you get
> an error, install it: in your `r-esfs` conda environment run
> `pip install leidenalg`. Alternatively replace with `sc$tl$louvain()`.

### Step 9b — Calculate ES matrices for Leiden secondary features

```{r calc-es-leiden}
# One-hot encoding of Leiden cluster labels (cells × n_clusters)
unique_leiden <- sort(unique(leiden_labels))

leiden_onehot <- Matrix::sparseMatrix(
  i    = seq_along(leiden_labels),
  j    = match(leiden_labels, unique_leiden),
  x    = 1,
  dims = c(length(leiden_labels), length(unique_leiden)),
  dimnames = list(colnames(obj), unique_leiden)
)

obj <- CalcESMatrices(
  obj,
  label                     = "Leiden_Clusts_Secondary_Features",
  save_matrices             = c("ESSs", "SGs"),
  secondary_features_matrix = leiden_onehot
)
```

### Step 9c — Run ES-CCF

For each gene, ES-CCF identifies which combination of Leiden clusters maximally
correlates with that gene's expression. The output (`Max_ESS_Features`) is
automatically cached in `misc` for the next step.

```{r run-esccf}
obj <- RunESCCF(
  obj,
  label   = "Leiden_Clusts_Secondary_Features",
  n_cores = -1L
)
# The message above shows the cache key to use in the next CalcESMatrices call
```

### Step 9d — Calculate ES matrices for Max_ESS_Features

The Max_ESS_Features matrix (cached by RunESCCF) is used as secondary features
to compute gene–cluster ESS scores. The matrix is restored automatically from `misc`.

```{r calc-es-max-ess}
obj <- CalcESMatrices(
  obj,
  label         = "Leiden_Clusts_Secondary_Features_Max_ESS_Features",
  save_matrices = c("ESSs")
)
```

### Step 9e — Run ES-FMG to select marker genes

ES-FMG uses simulated annealing to find a compact set of N genes that captures the
most non-redundant expression patterns across all identified cluster structures.

```{r run-esfmg}
fmg_result <- RunESFMG(
  obj,
  N           = 300L,
  label       = "Leiden_Clusts_Secondary_Features_Max_ESS_Features",
  num_reheats = 3L,
  resolution  = 1.0
)

chosen_genes <- fmg_result$genes
chosen_scores <- fmg_result$scores
obj <- fmg_result$object

message(sprintf("ES-FMG selected %d marker genes", length(chosen_genes)))
head(data.frame(gene = chosen_genes, score = chosen_scores))
```

### Step 9f — Visualise marker gene expression

```{r plot-markers}
# Choose a few genes to inspect — use chosen_genes for the full set
plot_genes <- c("PRSS3", "PDGFRA", "ATP6V0D2")

# Get the ES-GSS UMAP embedding as an R matrix for custom plotting
cell_umaps_embed <- GetESFSCellUMAPs(
  obj,
  gene_clust_labels = gene_clust_labels,
  top_ess_genes     = top_ess_genes,
  specific_cluster  = c(1L),
  n_neighbors       = 50L,
  min_dist          = 0.1,
  log_transformed   = TRUE,
  return_model      = FALSE
)
es_umap_embedding <- reticulate::py_to_r(cell_umaps_embed[[1]][[1]])  # cells × 2

for (gene in plot_genes) {
  raw_expr <- as.numeric(Seurat::LayerData(obj, layer = "counts")[gene, ])
  raw_expr <- pmin(raw_expr, quantile(raw_expr, 0.975))

  df_plot <- data.frame(
    UMAP1 = es_umap_embedding[, 1],
    UMAP2 = es_umap_embedding[, 2],
    Raw   = raw_expr
  )

  p_raw <- ggplot(df_plot, aes(UMAP1, UMAP2, color = Raw)) +
    geom_point(size = 0.5) +
    scale_color_gradient2(low = "blue", mid = "white", high = "red",
                          midpoint = median(raw_expr)) +
    theme_void() + labs(title = paste(gene, "— raw expression"), color = "Expr")

  print(p_raw)

  # If smoothed expression was extracted to Seurat, plot that too:
  if ("esfs.smoothed" %in% Seurat::Layers(obj)) {
    smooth_expr <- as.numeric(Seurat::LayerData(obj, layer = "esfs.smoothed")[gene, ])
    smooth_expr <- pmin(smooth_expr, quantile(smooth_expr, 0.975))
    df_plot$Smooth <- smooth_expr

    p_smooth <- ggplot(df_plot, aes(UMAP1, UMAP2, color = Smooth)) +
      geom_point(size = 0.5) +
      scale_color_gradient2(low = "blue", mid = "white", high = "red",
                            midpoint = median(smooth_expr)) +
      theme_void() + labs(title = paste(gene, "— kNN smoothed"), color = "Expr")

    print(p_smooth)
  }
}
```

---

## Step 10 — Save the UMAP model

Save the fitted UMAP model so new cells can be projected without re-running the
full workflow.

```{r save-umap-model}
# Re-run GetESFSCellUMAPs with return_model = TRUE to get the saveable model
cell_umaps_model <- GetESFSCellUMAPs(
  obj,
  gene_clust_labels = gene_clust_labels,
  top_ess_genes     = top_ess_genes,
  specific_cluster  = c(1L),
  n_neighbors       = 50L,
  min_dist          = 0.1,
  log_transformed   = TRUE,
  return_model      = TRUE
)

umap_model     <- cell_umaps_model[[1]][[1]]   # fitted umap-learn model
umap_gene_list <- reticulate::py_to_r(cell_umaps_model[[2]][[1]])  # gene names

# Retrieve the embedding coordinates from the fitted model
es_umap_embedding <- reticulate::py_to_r(umap_model$embedding_)   # cells × 2

SaveESFSUMAPModel(
  model           = umap_model,
  gene_list       = umap_gene_list,
  filepath        = file.path(path, "cluster1_umap.joblib"),
  log_transformed = TRUE
)
message("UMAP model saved to: ", file.path(path, "cluster1_umap.joblib"))
```

---

## Step 11 — Project query data onto the reference UMAP

Load the West et al. 2019 CTB/STB/EVT dataset and project it onto the early embryo
embedding using the saved UMAP model.

```{r load-umap-model}
bundle      <- LoadESFSUMAPModel(file.path(path, "cluster1_umap.joblib"))
model       <- bundle$model        # Python umap-learn model (supports .transform())
model_genes <- reticulate::py_to_r(bundle$gene_list)
```

```{r load-west2019}
# Load West2019 query data
west_sce <- zellkonverter::readH5AD(file.path(path, "West2019.h5ad"))
west_obj <- SeuratObject::as.Seurat(west_sce, counts = "X", data = NULL)

# Ensure unique cell names
colnames(west_obj) <- make.unique(colnames(west_obj))

message("West2019: ", ncol(west_obj), " cells × ", nrow(west_obj), " genes")
```

```{r project-west2019}
# Find gene overlap between reference gene set and query dataset
available_genes <- intersect(model_genes, rownames(west_obj))
missing_genes   <- setdiff(model_genes, rownames(west_obj))
message(sprintf("Available: %d genes | Missing (set to 0): %d genes",
                length(available_genes), length(missing_genes)))

# Build cells × genes expression matrix for the query dataset
west_counts <- t(Seurat::LayerData(west_obj, layer = "counts"))  # cells × genes

avail_mat <- west_counts[, available_genes, drop = FALSE]
if (length(missing_genes) > 0) {
  zero_mat  <- Matrix::Matrix(0,
    nrow = nrow(avail_mat), ncol = length(missing_genes),
    dimnames = list(rownames(avail_mat), missing_genes),
    sparse = TRUE)
  full_mat <- cbind(avail_mat, zero_mat)
} else {
  full_mat <- avail_mat
}
full_mat <- full_mat[, model_genes, drop = FALSE]  # reorder to match model gene order

# Log2(x+1) transform and project via the saved UMAP model
np        <- reticulate::import("numpy", convert = FALSE)
log2_mat  <- log2(as.matrix(full_mat) + 1)
west2019_overlay <- reticulate::py_to_r(
  model$transform(np$array(log2_mat, dtype = "float64"))
)
```

```{r plot-projection-timepoint}
df_ref  <- data.frame(UMAP1 = es_umap_embedding[, 1],
                      UMAP2 = es_umap_embedding[, 2])
df_west <- data.frame(UMAP1      = west2019_overlay[, 1],
                      UMAP2      = west2019_overlay[, 2],
                      Timepoint  = west_obj$Timepoints,
                      CellType   = west_obj$West_Labels)

ggplot() +
  geom_point(data = df_ref,  aes(UMAP1, UMAP2),
             color = "gainsboro", size = 0.5) +
  geom_point(data = df_west, aes(UMAP1, UMAP2, color = Timepoint),
             size = 1.5, alpha = 0.9) +
  theme_void() +
  labs(title = "West2019 — projected by timepoint", color = "Timepoint")
```

```{r plot-projection-celltype}
ggplot() +
  geom_point(data = df_ref,  aes(UMAP1, UMAP2),
             color = "gainsboro", size = 0.5) +
  geom_point(data = df_west, aes(UMAP1, UMAP2, color = CellType),
             size = 1.5, alpha = 0.9) +
  theme_void() +
  labs(title = "West2019 — projected by cell type", color = "Cell type")
```

---

## Save results

```{r save-results, eval=FALSE}
# Save the Seurat object with all ESFS results stored in misc
saveRDS(obj, file.path(path, "ES_seurat_obj.rds"))

# Save intermediate vectors for reproducibility
saveRDS(es_gss_genes,     file.path(path, "ES_GSS_Genes.rds"))
saveRDS(leiden_labels,    file.path(path, "Leiden_Clusts.rds"))
saveRDS(chosen_genes,     file.path(path, "ES_FMG_Marker_Genes.rds"))
saveRDS(es_umap_embedding, file.path(path, "ES_UMAP_Embedding.rds"))
```

---

## Session info

```{r session-info}
sessionInfo()
```
